---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rmdal
 
<!-- badges: start -->
<!-- badges: end -->

rmdal provides R bindings to the [MDAL (Mesh Data Abstraction Library)](https://www.mdal.xyz/), 
enabling R users to read and write unstructured mesh data from numerous formats used in 
hydrology, meteorology, and numerical modelling.

MDAL is an [OSGeo Community Project](https://www.osgeo.org/projects/mdal/) developed by 
[Lutra Consulting](https://www.lutraconsulting.co.uk/) and provides the mesh layer 
functionality in QGIS.

## Supported Formats

rmdal supports all formats provided by MDAL, including:
 
- **2DM** - SMS/TUFLOW mesh files
- **UGRID** - NetCDF unstructured grid (CF conventions)
- **Selafin** - TELEMAC hydrodynamic output
- **HEC-RAS 2D** - HEC-RAS hydraulic modelling
- **FLO-2D** - FLO-2D flood modelling
- **GRIB** - Meteorological data
- **NetCDF** - Generic scientific data
- **XMDF** - TUFLOW results
- And many more...

Use `mdal_drivers()` to see all available drivers and their capabilities.

## Installation

rmdal requires the MDAL library to be installed on your system.

### Install MDAL

**Via conda (recommended):**

```bash
conda install -c conda-forge mdal
```

**From source:**

```bash
git clone https://github.com/lutraconsulting/MDAL.git
cd MDAL && mkdir build && cd build
cmake -DCMAKE_BUILD_TYPE=Release ..
make && sudo make install
```

### Install rmdal

```r
# Install from GitHub
# install.packages("pak")
pak::pak("mdsumner/rmdal")
```

## Usage

```{r example, eval = FALSE}
library(rmdal)

# Check MDAL version
mdal_version()

# List available drivers
mdal_drivers()

# Load a mesh
mfile <- system.file("extdata/dem_with_holes/tnz.adf", package = "rmdal", mustWork = TRUE)
mesh <- mdal_load(mfile)

# Query mesh properties
mdal_mesh_vertex_count(mesh)
mdal_mesh_face_count(mesh)
mdal_mesh_projection(mesh)
```

```{r real-example, echo = FALSE, eval = TRUE}
library(rmdal)
mdal_version()
```

```{r drivers, eval = TRUE}
# Available drivers
mdal_drivers()[, c("name", "long_name", "can_read_mesh")]
```

## Roadmap

rmdal is under active development. Planned features include:

- [ ] Vertex, face, and edge geometry extraction
- [ ] Dataset group and dataset access (temporal data)
- [ ] S7 classes for mesh objects
- [ ] Integration with wk for geometry interchange
- [ ] Integration with rgl for 3D visualization
- [ ] Mesh creation and writing

## Data Model

MDAL provides access to unstructured mesh data with the following hierarchy:

```
Mesh
├── Vertices (x, y, z coordinates)
├── Faces (polygons defined by vertex indices)
├── Edges (1D elements for network meshes)
└── Dataset Groups (e.g., "Depth", "Velocity")
    └── Datasets (values at specific times)
        └── Values (scalar or vector, on vertices/faces/volumes)
```

Fidelity note: MDAL's data model is richer than what simple features (sf) can represent. 
Converting to sf necessarily loses information like temporal datasets, 3D volumes, and 
the vertex-index topology. rmdal aims to preserve MDAL's native model internally and 
provide targeted conversions for interoperability.

## Related Projects

- [MDAL](https://www.mdal.xyz/) - The underlying C++ library
- [QGIS](https://qgis.org/) - Uses MDAL for mesh layer support
- [Crayfish](https://plugins.qgis.org/plugins/crayfish/) - QGIS plugin for mesh visualization
- [anglr](https://github.com/hypertidy/anglr/) - R package Mesh creation and topology for spatial data
- [rgl](https://github.com/dmurdoch/rgl/) - R package 3D visualization system based on OpenGL.

## License

MIT © Michael Sumner

rmdal includes bindings to MDAL which is MIT licensed. 
Copyright © Lutra Consulting.
